---
output:
  github_document:
    html_preview: false
# used by altdoc
default-image-extension: ''
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# polars _(R Polars)_

<!-- TODO: add link to discord -->

<!-- badges: start -->
[![R-multiverse status](https://img.shields.io/badge/dynamic/json?url=https%3A%2F%2Fcommunity.r-multiverse.org%2Fapi%2Fpackages%2Fpolars&query=%24.Version&label=r-multiverse)](https://community.r-multiverse.org/polars)
[![R-universe status](https://rpolars.r-universe.dev/polars/badges/version)](https://rpolars.r-universe.dev/polars)
[![CRAN status](https://www.r-pkg.org/badges/version/polars)](https://CRAN.R-project.org/package=polars)
[![Docs dev version](https://img.shields.io/badge/docs-dev-blue.svg)](https://pola-rs.github.io/r-polars)
<!-- badges: end -->

Polars is a blazingly fast DataFrame library, written in Rust.

The goal of Polars is to deliver fast, efficient data processing that:

- Utilizes all available cores on your machine.
- Optimizes queries to reduce unneeded work/memory allocations.
- Handles datasets much larger than your available RAM.
- Follows a consistent and predictable API.
- Adheres to a strict schema (data-types should be known before running the query).

This polars R package provides the R bindings for Polars.
It can be used to process Polars DataFrames and other data structures,
convert objects between Polars and R, and can be integrated with other common R packages.

To learn more, read the [online documentation](https://pola-rs.github.io/r-polars/) for this R package,
and the [user guide](https://docs.pola.rs/) for Python/Rust Polars.

## Install

The recommended way to install this package is from the R-multiverse community repository:

```r
Sys.setenv(NOT_CRAN = "true")
install.packages("polars", repos = "https://community.r-multiverse.org")
```

More recent (development) version may be installed from the rpolars R-universe repository:

```r
Sys.setenv(NOT_CRAN = "true")
install.packages('polars', repos = c("https://rpolars.r-universe.dev", "https://cloud.r-project.org"))
```

<!-- TODO: link to the installation vignette -->

## Usage

To avoid conflicts with other packages and base R function names, many of polars'
functions are hosted in the `pl` environment, and accessible via the `pl$` prefix.
Most of the functions for polars objects should be chained with the `$` operator.

Additionally, the majority of the functions are intended to match the Python Polars API.

These mean that Polars queries written in Python and in R are very similar.

For example, writing the [example from the user guide of Python/Rust](https://docs.pola.rs/#example)
in R:

```{r}
library(polars)

# Prepare a CSV file
csv_file <- tempfile(fileext = ".csv")
write.csv(iris, csv_file, row.names = FALSE)

# Create a query plan (LazyFrame) with filtering and group aggregation
q <- pl$scan_csv(csv_file)$filter(pl$col("Sepal.Length") > 5)$group_by(
  "Species",
  .maintain_order = TRUE
)$agg(pl$all()$sum())

# Execute the query plan and collect the result as a Polars DataFrame
df <- q$collect()

df
```

There are also some functions to manipulate polars objects using base R and some popular other packages.

```{r}
# Subset a Polars DataFrame using the `[` operator
df[1:2, 1:2]

# Execute a query plan and collect the result as a tibble data frame
tibble::as_tibble(q)
```

```{r}
#| echo: false
unlink(csv_file)
```

The [Get Started vignette](https://pola-rs.github.io/r-polars/vignettes/polars.html)
(`vignette("polars")`) provides a more detailed introduction.

## Extensions

While one can use this package as-is, other packages build on it to
provide different APIs:

- [polarssql](https://rpolars.github.io/r-polarssql/) provides a polars
  backend for [DBI](https://dbi.r-dbi.org/) and [dbplyr](https://dbplyr.tidyverse.org/).
- [tidypolars](https://tidypolars.etiennebacher.com/) allows one to
  use the [tidyverse](https://www.tidyverse.org/) syntax while using the power of polars.

## Maintainers

- [SHIMA Tatsuya](https://github.com/eitsupi)
- [Etienne Bacher](https://github.com/etiennebacher)

[Version 0 of R Polars](https://github.com/rpolars/r-polars0) was previously maintained by
[Søren Havelund Welling](https://github.com/sorhawell).

## Acknowledgements

This package is based on the [Polars open source project](https://github.com/pola-rs/polars),
originally founded by [Ritchie Vink](https://github.com/ritchie46) and developed by many contributors.

## License

MIT @ polars authors
