---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# polars

<!-- badges: start -->
[![R-universe status badge](https://rpolars.r-universe.dev/badges/polars)](https://rpolars.r-universe.dev)
[![Dev R-CMD-check](https://github.com/pola-rs/r-polars/actions/workflows/check.yaml/badge.svg)](https://github.com/pola-rs/r-polars/actions/workflows/check.yaml)
[![Docs release](https://img.shields.io/badge/docs-release-blue.svg)](https://rpolars.github.io)
[![Docs dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://rpolars.github.io/dev)
<!-- badges: end -->

The **polars** package for R gives users access to [a lightning
fast](https://duckdblabs.github.io/db-benchmark/) Data Frame library written in
Rust. [Polars](https://www.pola.rs/)' embarrassingly parallel execution, cache
efficient algorithms and expressive API makes it perfect for efficient data
wrangling, data pipelines, snappy APIs and so much more. Polars also supports
"streaming mode" for out-of-memory operations. This allows users to analyze
datasets many times larger than RAM.

Documentation can be found on the **r-polars**
[homepage](https://rpolars.github.io).

The primary developer of the upstream Polars project is 
Ritchie Vink ([@ritchie46](https://github.com/ritchie46)). 
This R port is maintained by 
Søren Welling ([@sorhawell](https://github.com/sorhawell)) and
[contributors](https://github.com/pola-rs/r-polars/graphs/contributors).
Consider joining our [Discord](https://discord.gg/4UfP5cfBE7) (subchannel) for
additional help and discussion.


## Install

The package is not yet available on CRAN. But we provide convenient installation
options for a variety of operating systems:

### R-universe

[R-universe](https://rpolars.r-universe.dev/polars#install) provides
pre-compiled **polars** binaries for Windows and MacOS (x86_64), with source
builds for other platforms. Please see further below for binary install options
on Linux.

```r
install.packages("polars", repos = "https://rpolars.r-universe.dev")
```

Special thanks to Jeroen Ooms ([@jeroen](https://github.com/jeroen)) for the
excellent R-universe support.

### GitHub releases

We also provide pre-compiled binaries for various operating systems, as well as 
source installs, on our
[GitHub releases](https://github.com/pola-rs/r-polars/releases) page. You can
download and install these files manually, or install directly from R. Simply
match the URL for your operating system and the desired release. For example, to
install the latest release of **polars** on Linux (x86_64) one would use:

```r
install.packages(
  "https://github.com/pola-rs/r-polars/releases/latest/download/polars__x86_64-pc-linux-gnu.gz",
  repos = NULL 
)
```

Similarly for Windows
([URL](https://github.com/pola-rs/r-polars/releases/latest/download/polars.zip)) 
and MacOS (x86_64,
[URL](https://github.com/pola-rs/r-polars/releases/latest/download/polars__x86_64-apple-darwin17.0.tgz)).
Just remember to invoke the `repos = NULL` argument if you are installing these
binary builds directly from within R.

One exception worth noting is MacOS (arm64), i.e. systems based on the new M1/M2
"Silicon" chips. To install **polars** on one of these machines, we need to
build the package from source and this requires 
[Xcode](https://developer.apple.com/xcode/) (`xcode-select --install`). 
Once Xcode is installed, you can run the below code chunk to build
**polars**. The corresponding `Makevars` script will download a ~200MB
cross-compiled object file, while your machine links and builds the final R
package.

```r
# install.packages("remotes")

remotes::install_github(
  "https://github.com/pola-rs/r-polars",
  ref = "long_arms64", force =TRUE
)
```
 
Please [file an issue](https://github.com/pola-rs/r-polars/issues) if you
require a different target or operating system build. Finally, see the bottom of
this README for details on how to install rust to build from source (only
relevant for developers, or users of unsupported operating systems).


### r2u
Speeding up your workflow? On Ubuntu, install polars + arrow from binaries and resolve
system dependencies reliably and quickly with r2u ([see link for configuration](https://eddelbuettel.github.io/r2u/)).
```r
rp = c("https://rpolars.r-universe.dev/bin/linux/jammy/4.2", "https://cloud.r-project.org")
install.packages(c("polars", "arrow"), repos = rp)
```

## Quickstart example

The [Get Started](https://rpolars.github.io/articles/polars.html) vignette (`vignette("polars")`) contains a series of detailed
examples, but here is a quick illustration.

**polars** is a very powerful package with many functions. To avoid conflicts
with other packages and base R function names, **polars**'s top level functions
are hosted in the `pl` namespace, and accessible via the `pl$` prefix. To
convert an R data frame to a Polars `DataFrame`, we call:

```{r}
library(polars)

dat = pl$DataFrame(mtcars)
dat
```

This `DataFrame` object can be manipulated using many of the usual R functions and accessors, e.g.:

```{r}
dat[1:4, c("mpg", "qsec", "hp")]
```

However, the true power of Polars is unlocked by using *methods*, which are
encapsulated in the `DataFrame` object itself. For example, we can chain the
`$groupby()` and the `$mean()` methods to compute group-wise means for each
column of the dataset:

```{r}
dat$groupby("cyl")$mean()
```

[The **polars** vignette](https://rpolars.github.io/articles/polars.html)
contains many more examples of how to use the package to:

* Read CSV, JSON, Parquet, and other file formats.
* Filter rows and select columns.
* Modify and create new columns.
* Sort data.
* Join, group, and aggregate data.
* Join and concatenate datasets.
* Group and aggregate.
* Handle missing values.
* Work with dates and times.
* Use the lazy execution engine for maximum performance and memory-efficient operations.


## Development and Contributions

Contributions are very welcome!

As of March 2023, **polars** has now reached nearly 100% coverage of the
underlying "lazy" Expr syntax. While translation of the "eager" syntax is still
a little further behind, you should be able to do just about everything using
`$select()` + `$with_columns()`. Most of the methods associated with
`DataFrame` and `LazyFrame` classes have been implemented, but not all. There
is still much to do, and your help would be much appreciated!

If you spot missing functionality---implemented in Python but not
R---please let us know on Github.

### System dependencies

To install the development version of Polars or develop new features, you will
to install the Rust toolchain:

- Install [`rustup`](https://rustup.rs/), the cross-platform Rust installer. Then:
  ```sh
  rustup toolchain install nightly
  rustup default nightly
  ```
- Windows: Make sure the latest version of [Rtools](https://cran.r-project.org/bin/windows/Rtools/) is installed and on your PATH.
- MacOS: Make sure [`Xcode`](https://developer.apple.com/support/xcode/) is installed.
- Install [CMake](https://cmake.org/) and add it to your PATH.


### Implementing new features

Here are the steps required for an example contribution, where we are implementing the 
[cosine expression](https://rpolars.github.io/reference/Expr_cos.html):

 - Look up the [polars.Expr.cos method in py-polars documentation](https://pola-rs.github.io/polars/py-polars/html/reference/expressions/api/polars.Expr.cos.html).
 - Press the `[source]` button to see the [Python implementation](https://github.com/pola-rs/polars/blob/d23bbd2f14f1cd7ae2e27e1954a2dc4276501eef/py-polars/polars/expr/expr.py#L5892-L5914)
 - Find the cos [py-polars rust implementation](https://github.com/pola-rs/polars/blob/a1afbc4b78f5850314351f7e85ded95fd68b6453/py-polars/src/lazy/dsl.rs#L396) (likely just a simple call to the Rust-Polars API)
 - Adapt the Rust part and place it [here](https://github.com/pola-rs/r-polars/blob/c56c49a6fc172685f50c15fffe3d14231297ad97/src/rust/src/rdataframe/rexpr.rs#L754).
 - Adapt the Python frontend syntax to R and place it [here](https://github.com/pola-rs/r-polars/blob/c56c49a6fc172685f50c15fffe3d14231297ad97/R/expr__expr.R#L3138). Add the roxygen docs + examples above.
 - Notice we use `Expr_cos = "use_extendr_wrapper"`, it means we're just using unmodified the [extendr auto-generated wrapper](https://github.com/pola-rs/r-polars/blob/c56c49a6fc172685f50c15fffe3d14231297ad97/R/extendr-wrappers.R#L253)
 - Write a test [here](https://github.com/pola-rs/r-polars/blob/c56c49a6fc172685f50c15fffe3d14231297ad97/tests/testthat/test-expr.R#L1921).
 - Run `renv::restore()` and resolve all R packages
 - Run `rextendr::document()` to recompile and confirm the added method functions as intended, e.g. `pl$DataFrame(a=c(0,pi/2,pi,NA_real_))$select(pl$col("a")$cos())`
 - Run `devtools::test()`. See below for how to set up your development environment correctly.
 
Note that PRs to **polars** will be automatically be built and tested on all
platforms as part of our GitHub Actions workflow. A more detailed description of
the development environment and workflow for local builds is provided below.

 
### Development workflow

Assuming the system dependencies have been met (above), the typical **polars**
development workflow is as follows: 

**Step 1:** Fork the **polars** repo on GitHub and then clone it locally.

```sh
git clone git@github.com:<YOUR-GITHUB-ACCOUNT>/r-polars.git
cd r-polars
```

**Step 2:** Build the package and install the suggested package dependencies.

- Option A: Using **devtools**.

  ```r
  Rscript -e 'devtools::install(pkg = ".", dependencies = TRUE)' 
  ```

- Option B: Using **renv**.

  ```r
  # Rscript -e 'install.packages("renv")'
  Rscript -e 'renv::activate(); renv::restore()'
  ```

**Step 3:** Make your proposed changes to the R and/or Rust code. Don't forget to run: 

  ```r
  rextendr::document() # compile Rust code + update wrappers & docs
  devtools::test()     # run all unit tests
  ```

**Step 4 (optional):** Build the package locally.

  ```r
  R CMD INSTALL --no-multiarch --with-keep.source polars
  ```

**Step 5:** Commit your changes and submit a PR to the main **polars** repo. 

- As aside, notice that `./renv.lock` sets all R packages during the server build.

 
*Tip:* To speed up the local R CMD check, run the following: 

  ```r
  devtools::check(
    env_vars = list(RPOLARS_RUST_SOURCE="/YOUR/OWN/ABSOLUTE/PATH/r-polars/src/rust"),
    check_dir = "./check/"
    )
  source("./inst/misc/filter_rcmdcheck.R")
  Sys.sleep(5)
  unlink("check",recursive = TRUE, force =TRUE)
  ```

- The `RPOLARS_RUST_SOURCE` environment variable allows **polars** to recover the Cargo cache even if source files have been moved. Replace with your own absolute path to your local clone!
- `filter_rcmdcheck.R` removes known warnings from final check report.
- `unlink("check")` cleans up.
