---
output:
  github_document:
    html_preview: false
# used by altdoc
default-image-extension: ''
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# polars

<!-- badges: start -->
[![R-universe status badge](https://rpolars.r-universe.dev/badges/polars)](https://rpolars.r-universe.dev)
[![CRAN status](https://www.r-pkg.org/badges/version/polars)](https://CRAN.R-project.org/package=polars)
[![Dev R-CMD-check](https://github.com/pola-rs/r-polars/actions/workflows/check.yaml/badge.svg)](https://github.com/pola-rs/r-polars/actions/workflows/check.yaml)
[![Docs release](https://img.shields.io/badge/docs-release-blue.svg)](https://rpolars.github.io)
<!-- badges: end -->

The **polars** package for R gives users access to [a lightning
fast](https://duckdblabs.github.io/db-benchmark/) Data Frame library written in
Rust. [Polars](https://www.pola.rs/)' embarrassingly parallel execution, cache
efficient algorithms and expressive API makes it perfect for efficient data
wrangling, data pipelines, snappy APIs, and much more besides. Polars also supports
"streaming mode" for out-of-memory operations. This allows users to analyze
datasets many times larger than RAM.

Documentation can be found on the **r-polars**
[homepage](https://rpolars.github.io).

The primary developer of the upstream Polars project is
Ritchie Vink ([@ritchie46](https://github.com/ritchie46)).
This R port is maintained by
Søren Welling ([@sorhawell](https://github.com/sorhawell)) and
[contributors](https://github.com/pola-rs/r-polars/graphs/contributors).
Consider joining our [Discord](https://discord.com/invite/4UfP5cfBE7) (subchannel) for
additional help and discussion.

## Install

The package can be installed from R-universe, or GitHub.

Some platforms can install pre-compiled binaries, and others will need to build from source.

````{comment}

### CRAN

CRAN provides pre-compiled binaries for Windows (x86_64) and macOS.

Binary packages on CRAN are compiled by stable Rust, with nightly features disabled.

```r
install.packages("polars")
```

````

### R-universe (recommended)

[R-universe](https://rpolars.r-universe.dev/polars#install) provides
pre-compiled **polars** binaries for Windows (x86_64), macOS (x86_64) and Ubuntu 22.04 (x86_64)
with source builds for other platforms.

Binary packages on R-universe are compiled by nightly Rust, with nightly features enabled.

```r
# Binary installation for x86_64 Windows and macOS, source for other platforms
install.packages("polars", repos = "https://rpolars.r-universe.dev")
```

```r
# Binary installation for Ubuntu 22.04 (x86_64)
install.packages("polars", repos = "https://rpolars.r-universe.dev/bin/linux/jammy/4.3")
```

Special thanks to Jeroen Ooms ([@jeroen](https://github.com/jeroen)) for the
excellent R-universe support.

### GitHub releases

Binary packages on GitHub releases are compiled by nightly Rust, with nightly features enabled.

See latest and all previous [GitHub Releases here](https://github.com/pola-rs/r-polars/releases).

You can download and install these files manually, or install directly
from R. Simply match the URL for your operating system and the desired release. For example, to
install the latest release of **polars** on one can use:

Just remember to invoke the `repos = NULL` argument if you are installing these
binary builds directly from within R.

#### Linux (x86_64)

```r
install.packages(
  "https://github.com/pola-rs/r-polars/releases/latest/download/polars__x86_64-pc-linux-gnu.gz",
  repos = NULL
)
```

#### Windows (x86_64)

```r
install.packages(
  "https://github.com/pola-rs/r-polars/releases/latest/download/polars.zip",
  repos = NULL
)
```

#### macOS (x86_64)

```r
install.packages(
  "https://github.com/pola-rs/r-polars/releases/latest/download/polars__x86_64-apple-darwin20.tgz",
  repos = NULL
)
```

### Build from source

For source installation, pre-built Rust libraries may be available
if the environment variable `NOT_CRAN` is set to `"true"`. (Or, set `LIBR_POLARS_BUILD` to `"false"`)

```r
Sys.setenv(NOT_CRAN = "true")
install.packages("polars", repos = "https://rpolars.r-universe.dev")
```

Otherwise, the Rust library will be built from source.
the Rust toolchain (Rust `r RcppTOML::parseTOML("src/rust/Cargo.toml")$package$"rust-version"` or later) must be configured.

Please check the <https://github.com/r-rust/hellorust> repository for about Rust code in R packages.

```{r, include = FALSE}
rust_toolchain_version = read.dcf(
  "DESCRIPTION",
  fields = "Config/polars/RustToolchainVersion", all = TRUE
)[1, 1]
```

During source installation, some environment variables can be set to enable Rust features and profile changes.

- `RPOLARS_FULL_FEATURES="true"` (Build with nightly feature enabled, requires Rust toolchain `r rust_toolchain_version`)
- `RPOLARS_PROFILE="release-optimized"` (Build with more optimization)

## Quickstart example

The [Get Started](https://rpolars.github.io/articles/polars/) vignette (`vignette("polars")`) contains a series of detailed
examples, but here is a quick illustration.

**polars** is a very powerful package with many functions. To avoid conflicts
with other packages and base R function names, **polars**'s top level functions
are hosted in the `pl` namespace, and accessible via the `pl$` prefix. To
convert an R data frame to a Polars `DataFrame`, we call:

```{r}
library(polars)

dat = pl$DataFrame(mtcars)
dat
```

This `DataFrame` object can be manipulated using many of the usual R functions and accessors, e.g.:

```{r}
dat[1:4, c("mpg", "qsec", "hp")]
```

However, the true power of Polars is unlocked by using *methods*, which are
encapsulated in the `DataFrame` object itself. For example, we can chain the
`$group_by()` and the `$mean()` methods to compute group-wise means for each
column of the dataset:

```{r}
dat$group_by("cyl", maintain_order = TRUE)$mean()
```

Note that we use `maintain_order = TRUE` so that `polars` always keeps the groups
in the same order as they are in the original data.

[The **polars** vignette](https://rpolars.github.io/articles/polars/)
contains many more examples of how to use the package to:

* Read CSV, JSON, Parquet, and other file formats.
* Filter rows and select columns.
* Modify and create new columns.
* Group by and aggregate.
* Reshape data.
* Join and concatenate different datasets.
* Sort data.
* Work with dates and times.
* Handle missing values.
* Use the lazy execution engine for maximum performance and memory-efficient operations.
* Etc.
