---
title: "Introduction to rpolars"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to rpolars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## What is Polars?

[Polars](http://pola.rs) is among the
[fastest](https://h2oai.github.io/db-benchmark/) data table query 
libraries available in any language. While the underlying computation engine is
written in Rust, the original Polars bindings were implemented for Python. The
**rpolars** package provides equivalent bindings from R.

Polars syntax is similar to that of Spark, but the workflow is column-oriented
rather than row-oriented. Since R is itself a column-oriented language, this
should immediately feel familiar to most R users. Like Spark and modern SQL
variants, Polars optimizes queries for memory consumption and speed, so you
don't have to. However, unlike Spark, Polars is natively multithreaded instead
of multinode(d). This makes (r)polars much simpler to install and can be used as
one would any other R package.

All told, **rpolars** users can probably expect a 5-10 speedup compared to
**dplyr** for simple transformations on datasets >500Mb. The automatic Polars
optimization framework means that that this speed boost can be even greater for
complex queries that chain together many operations. Performance is similar to
that of **data.table**, although **rpolars** supports additional functionality
via its relationship to the Apache Arrow memory model. For example, it can scan
multiple Parquet files and datasets (e.g., directories with Hive-style
partitioning) and selectively import random subsets without having to read all
of the data. Furthermore, many operations can be performed efficiently
out-of-core, enabling users to work with larger than RAM data.

This R port relies on the excellent [**extendr**](https://github.com/extendr)
package, which is the R equivalent to pyo3+maturin. **extendr** is very
convenient for calling Rust from R, and vice versa, and is what we use to build
the **rpolars** package. Once built, however, **rpolars** has no other
dependencies other than R itself. This makes it very fast and lightweight to
install, and so **rpolars** can immediately be used to tackle your big (or
small!) data wrangling tasks.

The rest of this vignette contains some simple examples to get you going. Note
that the **rpolars** syntax and expressions are (deliberately) as close to the
Python implementation as possible, so you can always refer to the
[(py)polars book](https://pola-rs.github.io/polars-book/user-guide/) for more 
ideas. Just remember to switch out the `.` (Python) for `$` (R) when chaining 
methods. For example, here are two equivalent lines of code for some
hypothetical dataset.

```
# Python
df.select([pl.col(["A", "B"]).sum()]))

# R
df$select(pl$col(c("A", "B"))$sum())
```

## Constructors, object types and data types

Let's load the **rpolars** package and get our feet wet with a tongue-in-cheek
version of "Hello, world".

```{r}
library(rpolars)

pl$col("hello")$sum()$over("world", "from")$alias("polars")
```

The key point of this example is the following: All **rpolars** constructors can
be accessed via the `pl$` prefix. We'll see this repeatedly in the examples that
follow.

### Data types

Polars is
[strongly typed](https://en.wikipedia.org/wiki/Strong_and_weak_typing) and new
types can be created with the `dtypes` constructor. For example:

```{r}
# pl$dypes # list all types
pl$dtypes$Float64
```

The full list of valid Polars types can be found by typing `pl$dtypes` 
into your R console. These include _Boolean_, _Float32(64)_, _Int32(64)_,
_Utf8_, _Categorical_, _Date_, etc. Note that some type names differ from what
they are called in R (e.g., _Boolean_ in Polars is equivalent to `logical()` in
R). This might occasionally require you to look up a specific type. But the good
news is that **rpolars** generally does a good job of inferring types
automatically. We'll see some examples of type coercion later on.

### `Series` and `DataFrames`

The Polars analog to an R vector is a `Series`.

```{r}
pl$Series((1:5)*5, name = "my_series")
```

Recall that Polars and R are both column-orientated. So you can think of
`DataFrames` (data.frames) as being made up of a collection of `Series`
(vectors). In fact, you can create a new Polars `DataFrame` as a mix of `Series`
and/or regular R vectors.

```{r}
pl$DataFrame(
  pl$Series((1:5) * 5, "a"),
  pl$Series(letters[1:5], "b"),
  newname = pl$Series(c(1,2,3,4,5), "oldname"), # overwrite oldname with newname
  c(5,4,3,2,1), # unnamed vector
  named_vector = c(15,14,13,12,11) , # named provide
  c(5,4,3,2,0)
)
```

For the rest of this vignette, we'll focus on how to perform some common query
tasks on DataFrames.

## Common data tasks

_Method chaining_ is a core part of the **rpolars** workflow. If you're coming
from one of the other popular data wrangling libraries in R, then you probably
already have an innate sense of what this means. For instance,

- In **dplyr** we use a pipe operator, e.g. `dat |> filter(...) |> select(...)`
- In **data.table** we use its indexing syntax, e.g. `DT[i, j, by][...]`
- Etc.

In **rpolars** our method chaining syntax takes the form `object$m1()$m2()`,
where `object` is our data object, and `m1()` and `m2()` are appropriate 
[methods](https://rpolars.github.io/reference/index.html), like an subsetting
or aggregation expressions.

This might all seem a little abstract, so let's walk through some quick
examples to help make things concrete. Start by creating a polars "DataFrame"
object.^[Similar to how (most) **data.table** operations are limited to objects
of class `data.table`, we can only perform polars operations on objects that
have been converted to an appropriate **rpolars** class.] Later on, we'll see
how to do this automatically when importing data from disk. But for the moment,
it will suffice to coerce R's built-in `mtcars` dataset to a polars DataFrame.

```{r}
# Create a polars DataFrame from mtcars
dat = pl$DataFrame(mtcars)
dat
```

(As an aside, notice that **rpolars** DataFrames have their own "pretty" print
method.)

With our DataFrame in hand, we can now start chaining together various methods 
(expressions) to manipulate it in different ways. For example, we can subset the
data by rows
([`filter()`](https://rpolars.github.io/reference/filter-open-paren-close-paren.html))
and also columns
([`select()`](https://rpolars.github.io/reference/DataFrame_select.html)).

```{r}
dat$filter(
  pl$col("cyl")==6
)$select(
  pl$col(c("mpg", "hp", "cyl"))
)
```

The `select()` method also supports data modification, so you can simultaneously
transform it while you are subsetting. However, the result will exclude any
columns that weren't specified as part of the expression. To modify or add some 
columns---whilst preserving all others in the dataset---it is therefore better 
to use the
[`with_columns()`](https://rpolars.github.io/reference/DataFrame_with_columns.html)
method. This next code chunk is equivalent to 
`mtcars |> dplyr::mutate(sum_mpg=sum(mpg), sum_hp=sum(hp), .by = cyl)`.

```{r}
# Add the grouped sums of some selected columns.
dat$with_columns(
  pl$col("mpg")$sum()$over("cyl")$alias("sum_mpg"),
  pl$col("hp")$sum()$over("cyl")$alias("sum_hp")
)
```

For what it's worth, if we didn't care about overwriting the columns, the
previous query could have been written more concisely as:

```{r}
dat$with_columns(
  pl$col(c("mpg", "hp"))$sum()$over("cyl")
)
```

Similarly, here's how we could have aggregated (i.e., collapsed) the dataset
by groups instead of modifying them. We need simply invoke the `groupby()` and
[`agg()`](https://rpolars.github.io/reference/Expr_agg_groups.html) methods.

```{r}
dat$groupby(
  "cyl"
)$agg(
  pl$col(c("mpg", "hp"))$sum()$sort()
)
```

(The final `sort()` call is optional, since **rpolars** doesn't sort the results
of grouped operations by default. This is similar to what **data.table** does 
and is also true for newer versions of **dplyr**.)

The same principles of method chaining can be combined very flexibly to group by
multiple variables and aggregation types.

```{r}
dat$groupby(
  "cyl",
  pl$col("am")$cast(pl$Boolean)$alias("manual")
)$agg(
  pl$col("mpg")$mean()$alias("mean_mpg"),
  pl$col("hp")$median()$alias("med_hp")
)
```

As a final example of how **rpolars** can be used for standard data wrangling
tasks, let's implement a (left) join. For this example, we'll borrow some
datasets from the **nycflights13** package.

```{r}
data("flights", "planes", package = "nycflights13")
flights = pl$DataFrame(flights)
planes = pl$DataFrame(planes)

flights$join(
  planes, 
  on = "tailnum", 
  how = "left"
)
```

More information on the **rpolars** joining method can be found in the
[reference manual](https://rpolars.github.io/reference/DataFrame_join.html).

The package supports many other data manipulation operations, which we won't 
cover here. Hopefully, you will already have a sense of the key syntax features.
We now turn to another core idea of the Polars ecosytem: lazy execution.

## Lazy execution

While the "eager" execution engine of **rpolars** works perfectly well---as
evidenced by all of the previous examples---to get the most out of the package
you need to go _lazy_.
[Lazy execution](https://pola-rs.github.io/polars-book/user-guide/lazy-api/intro.html) 
enables several benefits, but the most important is that it improves
performance. Delaying execution until the last possible moment allows Polars to
apply automatic optimization to every query. Let's take a quick look.

To create a so-called "LazyFrame" from an existing object in memory, we can
invoke the `lazy()` constructor.

```{r}
ldat = dat$lazy()
ldat
```

Now consider what happens we run one our subsetting query from earlier on this
LazyFrame.

```{r}
subset_query = ldat$filter(
    pl$col("cyl")==6
)$select(
    pl$col(c("mpg", "hp", "cyl"))
)
  
subset_query
```

Right now we only have a tree of instructions. But underneath the hood, 
Polars has already worked out a more optimized version of the query. We can
view this optimized plan this by requesting it.

```{r}
subset_query$describe_optimized_plan()
```

Here we see a simple, but surprisingly effective component in query
optimization: _projection_. Changing the order in which our subsetting
operations occurs---in this case, subsetting on columns first---reduces the
memory overhead of the overall query and leads to a downstream speedup. Of
course, you would hardly notice difference for this small dataset. But the same
principles carry over to much bigger datasets and more complex queries.

To actually execute the plan, we just need tp invoke the `collect()` method.
This should feel very familiar if you have previously used other lazy execution
engines like those provided **arrow** or **dbplyr**.

```{r}
subset_query$collect()
```

## Data import

**rpolars** supports data import of both CSV and Parquet files formats. For
example:

```{r}
write.csv(airquality, "airquality.csv")

pl$read_csv("airquality.csv")
```

Again, however, the package works best if we take the lazy approach. 

```{r}
pl$lazy_csv_reader("airquality.csv")
```

We could obviously append a set of query operators to the above LazyFrame and
then collect the results. However, this workflow is even better suited to 
Parquet files and datasets, since we can leverage their efficient storage format 
on disk. Let's see an example.

```{r}
library(arrow)

dir.create("airquality-ds")
write_dataset(airquality, "airquality-ds", partitioning = "Month")
```

```{r}
# aq = read_parquet("airquality-ds/*/*.parquet) # eager version (okay)
aq = scan_parquet("airquality-ds/*/*.parquet")  # lazy version (better)

aq$filter(
  pl$col("Month")<=6
)$groupby(
  "Month"
)$agg(
  pl$col(c("Ozone", "Temp"))$mean()
)$collect()
```

We'll clean by removing our newly created files and directory before continuing.

```{r}
file.remove("airquality.csv")
unlink("airquality-ds", recursive = TRUE)
```

## Execute R functions within a Polars query

It is possible to mix R code with Polars by passing R functions to **rpolars**.
This can unlock a lot of flexibility, but note that it can inhibit performance.
R functions are typically slower, so we recommend using native Polars functions
and expressions wherever possible.

```{r}
pl$DataFrame(iris)$select(
  pl$col("Sepal.Length")$map(\(s) { # map with a R function
    x = s$to_r_vector() # convert from Polars Series to a native R vector
    x[x>=5] = 10
    x[1:10] # if return is R vector, it will automatically be converted to Polars Series again
    })
)$as_data_frame()
```

